---
title: "Analyzing Genre Influence on Spotify Track Characteristics"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

Necessary R libraries:
```{r, results='hide'}
library(base)
library(reticulate) #for Python interface
library(psych) #for skew/kurt functions
library(moments)
library(stats)
library(lsr)
```

Necessary Python libraries:
```{python, results='hide'}
import kaggle
import requests
import kaggle.cli
import sys
import numpy
import pandas as pd
from pathlib import Path
from zipfile import ZipFile
from kaggle.cli import main
```

```{python, results='hide'}
#download data set from Kaggle
dataset="thedevastator/spotify-tracks-genre-dataset"
sys.argv=[sys.argv[0]] + f"datasets download {dataset}".split(" ")

kaggle_zip=ZipFile(f"{dataset.split('/')[1]}.zip")
songdat={i.filename:pd.read_csv(kaggle_zip.open(i)) for i in kaggle_zip.infolist() }["train.csv"]

#Close the ZipFile object
kaggle_zip.close()
```

```{python, echo=FALSE}
import pandas as pd

songdat.drop(["Unnamed: 0"], axis=1, inplace=True) #Delete unused row
songdat.rename(columns={'track_genre': 'track_subgenre'}, inplace=True) #rename existing genre column to subgenre,(these genres are too specific IMO)

```

Clean data (python): group subgenres into broader catagories for new genre row
```{python}
import numpy

#Pop
songdat.loc[songdat['track_subgenre'].str.contains('pop', case=False, na=False)| songdat['track_subgenre'].str.contains('idol', case=False, na=False), 'track_genre']='Pop'

#Metal
songdat.loc[songdat['track_subgenre'].str.contains('metal', case=False, na=False) | (songdat['track_subgenre']=='hardcore'), 'track_genre']='Metal'

#Rock
rock_conditions=songdat['track_subgenre'].str.contains('rock', case=False, na=False) | songdat['track_subgenre'].isin(['goth', 'emo', 'garage', 'grunge', 'punk'])
songdat.loc[rock_conditions, 'track_genre']='Rock'

#Reggae
songdat.loc[songdat['track_subgenre'].str.contains('reggae', case=False, na=False) | (songdat['track_subgenre']=='ska'), 'track_genre']='Reggae'

#Electronic
electronic_conditions=songdat['track_subgenre'].str.contains('dance', case=False, na=False) | songdat['track_subgenre'].str.startswith('electro') | songdat['track_subgenre'].str.startswith('dub') | songdat['track_subgenre'].str.contains('house', case=False, na=False) | songdat['track_subgenre'].str.contains('techno', case=False, na=False) | songdat['track_subgenre'].isin(['edm', 'party', 'dubstep', 'hardstyle', 'breakbeat', 'club', 'drum-and-bass', 'idm', 'trance', 'trip-hop', 'industrial', 'happy'])
songdat.loc[electronic_conditions, 'track_genre']='Electronic'

#Disco
songdat.loc[songdat['track_subgenre'].str.contains('disco', case=False, na=False), 'track_genre']='Disco'

#Jazz
songdat.loc[songdat['track_subgenre'].str.contains('jazz', case=False, na=False), 'track_genre']='Jazz'

#Blues
songdat.loc[songdat['track_subgenre']=='blues', 'track_genre']='Blues'

#Folk
songdat.loc[songdat['track_subgenre'].isin(['folk', 'bluegrass']), 'track_genre']='Folk'

#Country
songdat.loc[songdat['track_subgenre']=='country', 'track_genre']='Country'

#R&B/Soul
songdat.loc[songdat['track_subgenre'].isin(['r-n-b', 'funk', 'soul', 'groove', 'afrobeat']), 'track_genre']='R&B/Soul'

#Latin
latin_conditions=songdat['track_subgenre'].str.contains('latin', case=False, na=False) | songdat['track_subgenre'].isin(['brazil', 'forro', 'mpb', 'pagode', 'salsa', 'samba', 'sertanejo', 'spanish'])
songdat.loc[latin_conditions, 'track_genre']='Latin'

#World
world_conditions=songdat['track_subgenre'].str.contains('world', case=False, na=False) | songdat['track_subgenre'].isin(['french', 'german', 'indian', 'malay', 'swedish', 'turkish', 'british', 'anime']) 
songdat.loc[world_conditions, 'track_genre']='World'

#Hip-hop
songdat.loc[songdat['track_subgenre'].isin(['hip-hop', 'sad']), 'track_genre']='Hip-hop'

#Kids
songdat.loc[songdat['track_subgenre'].isin(['children', 'kids', 'disney', 'comedy', 'show-tunes']), 'track_genre']='Kids/Family'

#Alternative
songdat.loc[songdat['track_subgenre'].isin(['alternative', 'chill', 'indie', 'new-age', 'sleep', 'ambient']), 'track_genre']='Alternative'

#Instrumental
songdat.loc[songdat['track_subgenre'].isin(['acoustic', 'guitar', 'piano']), 'track_genre']='Instrumental'

#Singer-songwriter
songdat.loc[songdat['track_subgenre'].str.contains('songwriter', case=False, na=False), 'track_genre']='Singer-songwriter'

#Classical
songdat.loc[songdat['track_subgenre'].isin(['classical', 'gospel', 'opera']), 'track_genre']='Classical'


#Capitalize first letter in Track Subgenre
songdat['track_subgenre']=songdat['track_subgenre'].apply(lambda x: x.capitalize())
```


Check if we've left behind any track_subgenre rows that haven't been matched with track_genre 

```{python}

if(songdat['track_genre'].isnull().sum())!=0:
    null_rows=songdat[songdat['track_genre'].isnull()]
    grouped_null=null_rows.groupby('track_subgenre').size().reset_index(name='count')
    print(grouped_null)


#Missed Grindcore, Honky-tonk, Iranian, Romance, Study, Tango

#Add Grindcore to Metal
songdat.loc[songdat['track_subgenre']=='Grindcore', 'track_genre']='Metal'

#Add Honky-tonk to Country
songdat.loc[songdat['track_subgenre']=='Honky-tonk', 'track_genre']='Country'

#Add Iranian to World
songdat.loc[songdat['track_subgenre']=='Iranian', 'track_genre']='World'

#Add Tango to Latin
songdat.loc[songdat['track_subgenre']=='Tango', 'track_genre']='Latin'

#Add Study to Alternative
songdat.loc[songdat['track_subgenre']=='Study', 'track_genre']='Alternative'

#Not sure where Romance should go, let's look at the artist names
romance=songdat[songdat['track_subgenre']=='Romance']
print(romance[['artists','track_name']].sort_values(by='artists'))
#Looks like Romance is a form of classical Russian music, add to World
songdat.loc[songdat['track_subgenre']=='Romance', 'track_genre']='World'

```



Choose numeric variables we want to compare against track_genre, our predictor variable. Let's try "popularity", "energy", "tempo", "duration_ms", "danceability","loudness". We'll scale all our raw values first to make comparison easier.

```{r}
# Convert py$songdat from Python to R dataframe


numeric_columns <- c("popularity", "energy", "tempo", "duration_ms", "danceability","loudness")

#Quick function to scale our rows quickly
scale_column <- function(df, column_name) {
  if (!column_name %in% colnames(df)) {
    stop("Column not found in the songdat")
  }
  
  df_scaled <- df
  df_scaled[[paste0(column_name, "_scaled")]] <- scale(df[[column_name]], scale=FALSE)
  
  return(df_scaled)
}

#cycle through all our rows in numeric_colummns, scaling each, and append to songdat
for (column_name in as.character(numeric_columns)) {
  scaled_column<-scale(py$songdat[column_name],scale=FALSE)
    py$songdat[[paste0(column_name, "_scaled")]] <-scaled_column[, 1]
}

```

Now check the Skew/Kurtosis values for all our dv's of interest.
```{r}

#Look at skew/kurt for scaled variables
skewness_scaled <- skew(py$songdat[paste0(numeric_columns,"_scaled")])
kurtosis_scaled <- kurtosis(py$songdat[paste0(numeric_columns,"_scaled")])


#Combine results into a data frame for easier viewing
skew_kurtosis_df <- data.frame(skewness=skewness_scaled, kurtosis=kurtosis_scaled)
print(skew_kurtosis_df)

#Duration_ms has way too much skew/kurt, 11.19503417 and 357.936795! Not worth attempting a transformation, too much variance
numeric_columns<-subset(numeric_columns, numeric_columns!="duration_ms") #Drop Duration_ms

```

All variables look good except for 2, duration_ms and loudness. We removed duration_ms because the skew/kurt was to extreme to fix through transformations. Loudness has a kurtosis of 8.9, which is problematic but might benefit from transformations. Let's try square root, inverse, and log transformations.

```{r}
song_pos=abs(py$songdat$loudness+1e-10)#There are 0s, to avoid NaN results add a really small number so it's technically positive

#Square Root
songNormSqt <- data.frame((song_pos+1)^0.5)
skewnessSqt <- skew(song_pos)
kurtosisSqt <- kurtosis(song_pos)

skewnessSqt
kurtosisSqt

#Inverse
songNormIn <- 1/(song_pos+1)
skewnessIn <- skew(songNormIn)
kurtosisIn <- kurtosis(songNormIn)

skewnessIn
kurtosisIn

#Log
songNormLg <- log10(song_pos+1)

skewnessLg <- skew(songNormLg)
kurtosisLg <- kurtosis(songNormLg)

kurtosisLg #Log looks good, kurtosis improved from 8.9 to 3.6 and skew is still minimal at .20.
skewnessLg 
```

The Log transformation for loudness looks great, skew dropped to 3.6 from 8.9. Let's apply this normally distributed version and use it as our loudness variable.
```{r}

#Let's replace our raw loudness data with our more normalized log transformation
#use unlist to ensure songNormLg is a 1-dim array
loudness_log_scaled_standardized <- scale(unlist(songNormLg)) #rescale

#assign scaled loudness values to songdat
py$songdat$loudness_log_scaled <- loudness_log_scaled_standardized[,1]

#update the numeric column list since we aren't using the raw
#Find the index of "loudness"
index_to_replace <- which(numeric_columns=="loudness")

#replace "loudness" with "loudness_log_scaled"
numeric_columns[index_to_replace] <- "loudness_log_scaled"
```

```{r, echo=FALSE, results='hide'}
rm(romance)
rm(column_name)
rm(index_to_replace)
rm(kurtosis_scaled)
rm(kurtosisIn)
rm(kurtosisLg)
rm(KurtosisSqt)
rm(skewness_scaled)
rm(skewnessIn)
rm(skewnessLg)
rm(skewnessSqt)
rm(song_pos)
rm(songNormIn)
gc()
```
Let's perform a series of one-way ANOVAs, using our numeric_column ("popularity", "energy", "tempo","danceability", "loudness_log_squared") as our dv's, and track_genre as our iv
```{r}

for(column_name in as.character(numeric_columns)){ 
print(column_name)
  aovTemp<-aov(py$songdat[[column_name]]~py$songdat$track_genre)
print(summary(aovTemp)) #Print ANOVA results 
print(etaSquared(aovTemp,type=3,anova=TRUE)) #...and eta-squard (effect size)
}
```

Surprisingly all of our dv's produced significant results, meaning track popularity, energy, tempo, danceability, and loudness are all influenced by a song's genre, p<.001. The largest effect sizes were seen for danceability (η2=.24) and loudness (η2=.22).

Since we know the means differ between track genre's, lets do some summary stats for each of our iv's by track genre. Pairwise analysis would be too much here since we have 18 distinct genre values, but we know we can trust our mean/median/modes due to significant ANOVA results.

```{r}
options(max.print =1000000)

#print the mean, median, sd for all our values of interest
summary_stats_by_genre <- aggregate(py$songdat[numeric_columns], by=list(track_genre=py$songdat$track_genre), FUN=function(x) c(mean=mean(x), median=median(x), sd=sd(x)))

print(summary_stats_by_genre)
```

Looks good, let's export our summary data into a CSV to import into Tableau!
```{r}
write.csv(summary_stats_by_genre, file="summary_stats_by_genre.csv", row.names=FALSE)
```

